% ============================================================
% SHA-256 as a mathematical object: a field guide and a
% research program.  (Baseline variant.)
%
% Build:  make            (all variants)
%         latexmk -pdf main.tex   (this variant only)
%
% Structure: preamble.tex holds all packages/macros; each section
% lives in sections/NN-name.tex; backmatter.tex holds the appendix
% and the clustered references. Sibling variants main-algebra.tex
% and main-projects.tex share everything but the front matter.
% ============================================================
\providecommand{\DocPdfTitle}{SHA-256 as a Mathematical Object}
\input{preamble}

\title{\SHA{} as a Mathematical Object:\\
       \large A Field Guide and a Program of Small Questions}
\author{[Author name here]\thanks{Working draft for mathematician
  colleagues. Comments very welcome.}}
\date{Last updated \lastupdated}

\begin{document}
\maketitle

\begin{abstract}
\noindent
The function \SHA{} maps arbitrary bit-strings of bounded length to
$256$-bit strings. It is one of the most heavily used mathematical objects
on Earth, and one of the least understood: a quarter-century of study has
produced neither a proof of any of its conjectured properties nor any
observable deviation from the behavior of a random function. This note
introduces \SHA{} historically and empirically---what it is for, why it has
lasted---and then frames it geometrically, as a self-map of the finite space
$(\Ztwo)^8$ built by deliberately alternating two incompatible group
structures on the same $2^{256}$-point set---a clash the $2$-adic metric
brings into sharp focus, sorting the machine operations by whether they
respect it and opening an unexpected line to the nonarchimedean dynamics
of $p$-adic disks. We describe the frameworks that
exist for analyzing such objects, why a large gap separates ``passes every
test we have'' from ``provably pseudorandom,'' and how the
complex-analytic tradition of Flajolet and collaborators, brought to bear
on the statistics of random mappings, offers concrete, measurable
coordinates for locating \SHA{} in the space of pseudorandom generators. We then open the adversary's own toolbox---%
differential cryptanalysis, the method that felled this function's
ancestors, with its remarkable secret history---and close with the view
from computational complexity: \textsc{sat} solvers, Impagliazzo's five
worlds, and what theorem provers did and did not prove. Throughout, we flag self-contained subprojects
suitable for undergraduates in short research windows.
\end{abstract}

\tableofcontents
\bigskip

\input{sections/00-howtoread}

\input{sections/01-josephus}
\input{sections/02-history-uses}
\input{sections/03-geometry}
\input{sections/04-random-maps}
\input{sections/045-iterated}
\input{sections/05-frameworks}
\input{sections/052-differential}
\input{sections/055-computability}
\input{sections/06-projects}

\input{backmatter}

\end{document}
