% ------------------------------------------------------------
% §B  Appendix: the w=8 centralizer census
% ------------------------------------------------------------
\section{Reference implementation: the centralizer census}
\label{app:censuscode}

The listing below (\texttt{tools/centralizer\_census.py} in the
distribution accompanying this document) is the computational companion
to \cref{sec:statespace}: it produces every number in \cref{tab:census}
and the surrounding text---the three-regime sorting of the narrowed
components, the random baselines that calibrate it, and the
Anashin-style ablation in which a single rotation collapses a
$2$-adically ergodic map onto random-permutation statistics. As with
\cref{app:sigmacode}, the point of printing it in full is
epistemological: every claim it supports becomes an artifact anyone
can audit and rerun (about ten seconds), rather than an argument to be
trusted.

The script has five movements, signposted in its comments.

First, the \emph{narrowed components}: the standard's diffusion and
chooser maps at word width $w = 8$ (rotation amounts reduced mod $8$,
round constants cut to their leading bytes), together with the bare
odometer $x \shaplus \mathtt{0xb5}$ and its one- and two-fold ARX
alternations. The selection is not neutral---it is chosen so that each
of the three regimes predicted in \cref{sec:statespace} (linear:
over-symmetric; odometer: under-symmetric; alternation: generic) is
represented and can confirm or embarrass the prediction.

Second, the \emph{exact counting machinery}. For a permutation $f$,
both the centralizer of $f$ in $\mathrm{Sym}(256)$ and the number of
arbitrary self-maps commuting with $f$ are closed-form functions of
the cycle type (the latter by a two-line argument recorded in a
docstring: a commuting $g$ is free on one point per cycle, and that
image point must be periodic of compatible period). For a
non-bijection no closed form exists, and the script instead counts
endomorphisms of the functional graph by dynamic programming over its
cycle-and-trees anatomy---enumeration is hopeless, since the answers
run to $10^{140}$. The counter self-tests against brute-force
enumeration on small domains before any census is reported.

Third, the \emph{structured classes} of \cref{tab:census}---the
translations $x \shaplus c$, the xor-translations $x \shaxor c$, and
the affine-circulant maps $x \mapsto L(x) \shaxor c$---checked
exhaustively. These are the symmetries an analyst could actually
exploit, each transparent in one of the state space's two rival
geometries; a nontrivial commuting member would be exactly the kind of
symmetry that, one regularity class up, Lubin's rigidity theorem
(\cref{thm:lubin}) says cannot exist without algebraic cause.

Fourth, the \emph{baselines}: $10^{4}$ random permutations and
$10^{2}$ random maps, which convert raw counts into percentiles---and
the calibrating observation, verified rather than assumed, that a
generic permutation's only structured symmetries are the constant maps
sitting at its Poisson-distributed fixed points.

Fifth, the \emph{ablation}: Anashin's transitivity mod $2^{k}$
\citep{Anashin2006} verified through $k = 16$ for the odometer and for
the Klimov--Shamir T-function $x \shaplus (x^{2} \vee
5)$~\citep{KlimovShamir2003}, then broken with one rotation and the
wreckage measured in the random-mapping coordinates of
\cref{sec:randommaps}~\citep{FlajoletOdlyzko1990}---down to the
frozen-bit ``scar'' by which the $1$-Lipschitz substrate stays visible
even after ablation.

\medskip
\lstinputlisting[language=Python]{tools/centralizer_census.py}
